In Class Assignment 3
=====================

Custom functions
----------------

We've encountered a number of functions before, from ```log()``` to ```class()``` to ```seq()``` to ```sort()``` - these are built-in functions
You can also build your own functions too - like the one below that takes a number and doubles it:
```{r}
times2 = function(x) {
  y = x * 2
  return(y)
}
```



Now you can call it and give it the argument you want to double
```{r}
times2(5)
times2(10)
```



The way to define a function is ```FUNCTION.NAME = function(ARGS) {FUNCTION.CODE}```

* FUNCTION.NAME is just any variable name
* ARGS is any number of arguments you want to pass to the function (here we just used one - x)
* FUNCTION.CODE is all the stuff you want the function to do - note that it must be within curly brackets

Let's break it down even further, by starting with the function call.
When you enter times2(10), it runs the code in times2 and sets the variable x to 10.
It knows that x is what you want to set because we defined x in the arguments: ```times2 = function(x) {...}```

It then starts running the code - the first thing it does is assign x * 2 to the variable y.
Note that y is a temporary variable that only exists while the function is running.
Since the function has now stopped, if you try to see what is in y, you'll get an error.


The last line of the function tells it to give you the value that is stored in y.
The return() command ends the function and returns whatever the stuff in the parentheses evaluates to.
This is your way of getting information out of the function and back to the console, since all variables are temporary.

Let's try another trivial example that adds two items:
```{r}
add = function(x,y) {
  return(x + y)
}

add(1,2)
add(5,5)
```

This function takes two arguments (x & y).
You can define multiple arguments by putting commas between them as above.
Also, note that return() first evaluates x + y, then gives it back to you.

Since functions return a value, you can call functions on the output from other functions.
For instance, the code below gives 10: 2*(4+1)
```{r}
times2(add(4,1))
```


You can view function code by typing in the function name without any arguments
```{r}
times2
```

This also works with some built-in functions, in case you're curious, like the function ```ls```

```{r}
ls
```

However, some built-ins are 'primitives' - they aren't written in R code so you can't see their guts.
You will know a function is a primitive if there is .Primitive() when you try to view the function.
```{r}
sum
```

Functions Self Assessment
-------------------------

Write a function, named ```first```, that takes a numeric vector as its only argument and returns the first element of the vector.

For instance, after running ```x <- first(c(1, 3, 5))```, ```x``` should be 1.
```{r}
# Your Code Here
```

Write a function named ```make_awesome``` that takes a string and adds " is Awesome!" to it.

Example:
```
make_awesome("Everything") => "Everything is Awesome!"
```
Note, the ```paste``` function may be useful.
```{r}
# Your Code Here
```

Getting Help
------------

Now you have enough R knowledge to do a lot, but you still might get stuck trying to figure out a function.
In that case, you don't need to panic, you just need the help() function.
You can call help() on any built-in function and an explanation will show up in the 'Help' window.
```{r}
help(sum)
```

This tells you what arguments the function takes, and what it does

You can also write this shorthand by starting with ?
```{r}
?sum
```

That works if you know the function but don't know how it works.
But in many cases, you won't even know what you're looking for.
For those times, there's ```??```.
This searches through all functions and gives you a list of functions that have help pages containing your search.

For instance, if you want to know how to do trigonometry:
```{r}
??trigonometry
```

From that you see base::Trig is a package - this means that Trig is part of the base R functions.
So we move on:
```{r}
?Trig
```
And now we know all about cos, sin, tan, acos, asin, atan, and atan2 - huzzah!


Matrices and Lists
------------------

We've learned how to store multiple pieces of data in vectors, but what if we want to structure the data into two dimensions...
Or store multiple types of data in one variable?
For this we have matrices and lists.

Matrices are just 2-dimensional vectors. The way to make one is the function: ```matrix(data,nrow,ncol)```.

* Data is a vector of values that gets read off into the columns, one by one
* nrow and ncol tell R how to shape the matrix

For instance, you can create a matrix with the numbers 1 to 6 and 2 rows, 3 columns with:
```{r}
matrix(1:6,nrow=2,ncol=3)
```

Note how the numbers count down the columns.
This is also different from the matrix with the same data, but 3 rows and 2 columns:
```{r}
matrix(1:6,nrow=3,ncol=2)
```

If you want the data to read across the rows instead of down the columns, set 'byrow' to TRUE
```{r}
matrix(1:6,nrow=3,ncol=2,byrow=TRUE)
```


Note that the length of the vector must have a length equal to nrow * ncol.
If not, can you guess what happens?
(Hint: it starts with 'r', ends with 'ecycling', and is bad)

Now lets store a matrix from 1-9
```{r}
junk.mat = matrix(1:9,nrow=3,ncol=3,byrow=TRUE)
junk.mat
```

We can access individual matrix items by index, similar to the way we do for vectors.
However, rather than one index, you should use two: [row,col]
```{r}
junk.mat[2,1] # Should be 4; second row, first column
junk.mat[3,2] # Should be 8; third row, second column
```



Accessing by individual indices also work (it flattens the matrix back into a vector), but can get confusing.
You can also take matrix slices by ranges, just like for vectors
```{r}
junk.mat[1:2,2:3] # Gets the upper right square
```

And you can also access individual rows or columns, using the index [row#,] or [,col#]
```{r}
junk.mat[1,] # Gives you the first row
junk.mat[,3] # Gives you the third column
junk.mat[1:2,] # Gives you rows 1 and 2
```

You can assign back to a matrix by index

```{r}
junk.mat[1,3] = 10 # Change one item
junk.mat
junk.mat[,2] = c(11,12,13) # Change a whole column
junk.mat
```

Finally, let's say you also want to find the shape of the matrix.
We can't just use the length command... it works, but only gives us the total number of items in the matrix
```{r}
length(junk.mat)
```


Instead, we need to use the dim() command.
This gives us a vector of two numbers... first the number of rows, then the number of columns
```{r}
dim(junk.mat)
```

Lists are a way of storing multiple vectors or data types into one data structure.
Remember, if you try to store numbers and characters together in a vector, it changes the numbers to characters.
```{r}
c(1,"A",2)
```

Likewise, you can't store vectors within a vector:
```{r}
c(1,c(2,3),4)
```

But with lists you can do both of these things
```{r}
list(1,"A",2)
list(1,c(2,3),4)
```

These look a little funny... that's because a list stores a number of different vectors.
The vector at [[1]] above is just 1, the vector at [[2]] is 2,3, etc.
Let's store that into a variable
```{r}
junk.list = list(1,c(2,3),4)
```

Accessing items out of a list is a little different - you must use double brackets if you want the data
```{r}
junk.list[[1]]
junk.list[[2]]
```

If you use single brackets, it just gives you a list of length 1 back - which you can't use as a number
```{r}
junk.list[1]
#junk.list[1] + 2 # ERROR!
junk.list[[1]] + 2 # Okay!
class(junk.list[[1]]) # A numeric - we pulled out the vector
class(junk.list[1])   # A list - we just got that element of the list
```

We can also create names lists within the list function:
```{r}
junk.list = list('A' = 1, 'B' = c(2,3), 'C' = 4)
junk.list
names(junk.list)
```

We can now access items from that list using the name
```{r}
junk.list[['A']]
```

But there's also a shortcut to the names using $ rather than the double brackets:
```{r}
junk.list$A
```

We can also change members of a list, or even elements of list members:
```{r}
junk.list$A = c(3,5) # Replaced element A with a vector
junk.list

junk.list[3] = 10 # Replaced the third element (C) with 10
junk.list

junk.list$B[1] = 12 # Replaced the first number in B with 12
junk.list

junk.list[['B']][2] = 16 # Replaced the second number in B with 16
junk.list
```


You can also add new items to a list by assigning to a new name within the list
```{r}
junk.list$newitem = c('Hi','there')
junk.list
```

Or remove items from a list by assigning NULL to an existing name.
NULL is a special variable that tells R that nothing is there
```{r}
junk.list$B = NULL
junk.list
```

Matrices and Lists Self Assessment
----------------------------------
Please describe the differences between the following operators and when they would be used:

1) Single, square brackets ```[]```
2) Double, square brackets ```[[]]```
3) Dollar sign ```$```
4) Quotation marks ```""```
5) Parentheses ```()```

Make a matrix that is a multiplication table up to four
```{r}
#Your Code Here
```

Modify the original matrix to remove the one's row and column.
```{r}
#Your Code Here
```

Make a list of classes your are in, named ```current_classes```, and a list of classes
you want to take in the Spring, named ```future_classes```.
```{r}
#Your Code Here
```

Make a list of lists composed of the two previous lists, named ```classes```. 
Use list names so that this line would work: ```classes$current_classes[[1]] ```
```{r}
#Your Code Here
```

Data frames
-----------

Now we move on to data frames - this is the way R stores most datasets.
Effectively, it is a list of vectors, but all of the vectors must be of equal length.
You can think of these as storing one data observation per row.
Each different type of information you have on those observations make up the columns.
Let's look at a builtin data frame - mtcars - as an example.
```{r}
mtcars
```

Each of the rows is named with the car type. 
Then each of the columns has different information about those cars (avg mpg, # of cylinders, etc.).
This data is rectangular - the number of rows is equal for every column

You can build a data frame just like a list:
```{r}
data.frame('Name' = c('A','B','C'), 'Num' = c(1,2,3), 'Logic' = c(TRUE,FALSE,TRUE))
```

Note how much easier this is to look at than a list with the same data.
It's easy to tell in a data frame which records go together - not so much in a list
```{r}
list('Name' = c('A','B','C'), 'Num' = c(1,2,3), 'Logic' = c(TRUE,FALSE,TRUE))
```

But note that if the vectors aren't equal length, R doesn' like it:
```{r}
#data.frame('Name' = c('A','B'), 'Num' = c(1,2,3), 'Logic' = c(TRUE,FALSE,TRUE)) # Note Name has 2 items only
```

And if you have a rectangular list, you can transform it into a data frame easily
```{r}
l = list('Name' = c('A','B','C'), 'Num' = c(1,2,3), 'Logic' = c(TRUE,FALSE,TRUE))
data.frame(l)
```

So let's store that data frame:
```{r}
junk.df = data.frame(l)
```

In addition to typing the name of a data frame in the console, RStudio let's us look at them in a window.
Remeber, you can click on the data frame in the Workspace tab, or use the ```View()``` command

We can also access data just like we did for lists:
```{r}
junk.df$Num
junk.df[[3]]
```

Or individual items from the vectors:
```{r}
junk.df$Num[1] # 1: the first element of Num
```

Because all the vectors are the same length, we can also use matrix indices to grab data.
However, it's advised not to use this - it's hard to tell exactly what you're getting back
```{r}
junk.df[1,2] # 1: the first row and second column (in this case Num)
```

Finally, we can access individual rows using matrix indices:
```{r}
junk.df[1,] # The first data observation
junk.df[2:3,] # The second and third data observation
```

We can also create data frames with row names by giving it the row.names argument
```{r}
junk.df2 = data.frame('Name' = c('A','B','C'), 'Num' = c(1,2,3), 'Logic' = c(TRUE,FALSE,TRUE),
                      row.names = c('Elem.1','Elem.2','Elem.3'))
junk.df2
```

Now we can access data by names:
```{r}
junk.df2['Elem.1','Num']
```

Or observations by name
```{r}
junk.df2['Elem.1',]
```

We can get these names back by using the ```dimnames()``` command.
This provides a list of names for each dimension
```{r}
dimnames(junk.df2)
```

If we use names(), we just get the column names
```{r}
names(junk.df2)
```

We can also set the dimnames just like we would a list:
```{r}
dimnames(junk.df2) = list(c('E1','E2','E3'),c('Name','Number','Logic'))
junk.df2
```

Or just the rownames, by changing the second item in the list:
```{r}
dimnames(junk.df2)[[1]] = c('El1','El2','El3')
```

Can also add new data to data frames, just by assigning it to an empty column
```{r}
junk.df2$NewCol = c(10,12,14)
junk.df2
```

And can add transformed data back to the table:
```{r}
junk.df2$Times2 = junk.df2$Number * 2
junk.df2
```

If you want to access only subsets of the observations based on conditionals, there are two ways of doing this.
First, you can have the conditional in the first part of the index.
```{r}
junk.df2[junk.df2$Logic == TRUE,]
```

Or you can just use the ```subset()``` command.
This takes two arguments - the data frame and the conditional.
Note that you don't need to tell ```subset()``` which data frame you are drawing the columns for.
It assumes that you are using the same data frame you are subsetting.
```{r}
subset(junk.df2,Logic == TRUE)
```

And just like conditionals from simulation, you can use multiple conditionals
```{r}
subset(junk.df2,Logic == TRUE & Number == 3)
```

This is very useful when you want to:

1) Review/analyze only parts of your data (e.g., only subjects in Condition 1)
2) Eliminate outliers or inappropriate data points

In either case, you can store the reduced table into another variable for easy access

```{r}
junk.df2.reduced = subset(junk.df2, Logic == TRUE)
junk.df2.reduced
```

The final bit of table access we'll go through involve the ```with()``` and ```attach()``` commands.

Sometimes you'll be accessing a number of variables in the same data frame at once.
It can get long and tedious to tell R that you are accessing the same data frame over and over.
So the with() command tells R that it should just assume the variables you call or from that data frame.
This takes two arguments: the data frame, and the expression you want to run.

For instance, this adds the Number data to the NewCol data

```{r}
with(junk.df2, Number + NewCol)
```

This is identical to typing:
```{r}
junk.df2$Number + junk.df2$NewCol
```

This gets very useful when you want to access only a subset of the table:
```{r}
with(subset(junk.df2,Logic == TRUE), Number + NewCol)
```

And then there are other times that you'll be working with one data frame exclusively.
In this case, you might be using those variables on every line, but don't want to write with() over and over.
For this, there's the attach command, which 'attaches' all of the data from the data frame to the console.
```{r}
attach(junk.df2)
```

Once you've attached a data frame, you can access the data without having to tell R the frame
```{r}
Number
NewCol
Number + NewCol
```

However, be careful - if you change any of the data, it will create a copy and won't affect the data frame
```{r}
Number[2] = 10
Number
junk.df2$Number
```

Note that 'Number' now shows up in your workspace

Finally, when you're done with the data frame, you can detach() it.
This will prevent default access to the data, and is good clean-up practice.
```{r}
detach(junk.df2)
#NewCol #This will give an error as it is detached
```

But any copied/changed data will remain
```{r}
Number
```

A final note on data frames... sometimes you want to know what type of data is stored in each column.
For this, you can use the ```str()``` command
```{r}
str(junk.df2)
```

Note that this tells you what each collumn is (factor, numeric, logical, etc.) and the first few observations.
This can be very useful if you're given a new dataset and want to know what it contains.

You might also want to get just the first few records.
For this you want to use the head() command - for the first six.
This won't have an effect with junk.df2 (there are only 3 records).
But look at mtcars
```{r}
mtcars
head(mtcars)
```

It's much easier to get a quick readout with 'head'

Dataframes Self Assessment
--------------------------
Show me the rows of ```mtcars``` that have cars with over 20 mpg (miles per gallon) and more that 4 cyl(inders).



Input/Output
============

In many cases you'll want to read and write data from files.
R makes this easy for you with the read.table() and write.table() commands.

To open up a file, first you need to make sure you're in the correct 'working directory'.
This tells the computer where to look for the files.
To learn where your working directory is set, you can use the ```getwd()``` command.
```{r}
getwd()
```

You can then use the setwd() command with the path argument to set it to where you want to go.
But there are much easier ways of doing this in RStudio.

First, you can use the Files tab.
Just go to the directory you want to access, then click 'More' and select 'Set As Working Directory'.
If you can see the file you want to open in the Files tab, then you're in the right place.

Next, you can select it manually.
Go to the 'Session' menu, select 'Set Working Directory' and 'Choose Directory'.
You will be able to search through your files to find the right directory.

Finally, if you are working on RScript and your data is in the same directory, this is easy.
Just go to the 'Session' menu, select 'Set Working Directory' and 'To Source File Location'.

Let's start with the F&D_Data.csv file
Download it then click on it in the Files pane
You'll see it's just some text that looks like:

    A,20,1.5
    A,30,0.5
    B,40,2
    B,50,5.5

This is a typical csv file - different columns of data separated by commas, with each observation on a new line.
You can read this in with the read.table() command
Note that you must set the 'sep' argument to ',' to tell it that commas count as data breaks
```{r}
read.table('F&D_Data.csv',sep = ',')
```

The columns aren't named - but you can do this with the col.names argument
```{r}
read.table('F&D_Data.csv',sep = ',',col.names = c('Cond','N1','N2'))
```

And can name the rows using the row.names argument
```{r}
read.table('F&D_Data.csv',sep = ',',col.names = c('Cond','N1','N2'),row.names = c('E1','E2','E3','E4'))
```

We can also select only some of the rows.
If you want a certain number of rows, you can use the nrows argument to tell R how many observations you want.
It will only read in the first n rows of the file.
```{r}
read.table('F&D_Data.csv',sep = ',',col.names = c('Cond','N1','N2'),nrows = 2)
```

Now let's take a look at the 'F&D_Data_v2.csv' file.
This is exactly like the Data1 file, but column names are already there.
If you just try to read in the table, it will think that the headers are part of the data. 
(Note: recent versions of R are sometimes smart enough to recognize headers,
but you shouldn't depend on this.)
```{r}
read.table('F&D_Data_v2.csv',sep = ',')
```

Instead, we want to use the 'header' argument to tell R that the first line is actually a header.
This sets the column names to whatever is in the first line.
```{r}
read.table('F&D_Data_v2.csv',sep=',',header = TRUE)
```


A shortcut to all of this is also the ```read.csv()``` command.
This is exactly like the read.table command, except it assumes that ```sep=','``` and ```header=TRUE```.
```{r}
read.csv('F&D_Data_v2.csv')
```

So now let's store the data:

```{r}
sample.data = read.csv('F&D_Data_v2.csv')
sample.data
```

It's then a data frame - we can do anything we want to it within R now.
```{r}
sample.data$N3 = sample.data$N1 * 4
sample.data
```


But when we're done, we might want to write our data frame back out to a file.
For this, we have the write.table() command.
This needs two arguments, the data frame and the file name.
```{r}
write.table(sample.data,'F&D_Data_Output.csv')
```


But open it up - this looks a bit messy
First, we didn't give is a sep argument, so it assumed spaces should separate columns.
```{r}
write.table(sample.data,'F&D_Data_Output.csv',sep=',')
```

If you open it up again, you'll note that the original data has been overwritten.
We can change this using by setting the append argument to TRUE.
This tells R not to overwrite the file, but instead to write to the end of the file.
```{r}
write.table(sample.data,'F&D_Data_Output.csv',sep=',',append=TRUE)
```

However, you'll see a warning when you do that - it tells you column names are appended.
Usually if you're just adding data, you don't want to rewrite the collumn names.
You can turn this off by setting col.names to FALSE.
```{r}
write.table(sample.data,'F&D_Data_Output.csv',sep=',',append=TRUE, col.names = FALSE)
```

Let's clear all of this and just get the basic data into this file:

```{r}
write.table(sample.data,'F&D_Data_Output.csv',sep=',')
```

Also, like ```read.csv``` was shorthand for ```read.table```, there is a ```write.csv``` command.
This is almost exactly like setting sep=',' in write.table
```{r}
write.csv(sample.data,'F&D_Data_Output.csv')
```

But open these two files side-by-side and you'll notice one important difference.
There is an extra "", at the start of the write.csv file.
If you tried opening the first output in Excel, the collumn names would be mismatched with the data.
It would put 'Cond' above the element names.
With that extra bit from write.csv, the element names header will be blank and everything else will match.

Finally, we can turn off writing the element names by setting row.names to FALSE
```{r}
write.csv(sample.data,'F&D_Data_Output.csv',row.names=FALSE)
```

Input/Output Self Assessment
----------------------------

Scan through the documentation for ```read.table```. What does the ```skip``` argument do and when might you use it?



Factors
-------

In the first lesson, we talked about different types of singluar data that could be stored in variables.
These could be numbers, character strings, or logical values.
Now it's time to learn about another type of data R uses a lot: factors.
Factors come up often when importing data tables.
For instance, what happens if we look at the class of 'Cond' in sample.data?.
```{r}
class(sample.data$Cond)
```

Notice it says "factor".
If you look in the variable, it looks slightly different than a character string...
Note the 'Levels'.
```{r}
sample.data$Cond
```

So why use factors instead of characters?
Factors are a way for R to store different conditions in a way that's easy to use and save space.
For instance, let's say you're running an experiment with three drugs, a control group, and 20 subjects.
You could store these conditions a character strings.
```{r}
cond.char = rep(c('DrugA','DrugB','DrugC','Control'),each = 5)
```

Okay... that works and differentiates the conditions, but it takes up a lot of space.
Computers need more memory to store character strings than numbers.
This isn't really a problem with 20 subjects, but some datasets can have millions of records.
That space would add up fast.

So how about we store them as a number: 1,2,3,4?
```{r}
cond.num = rep(c(1,2,3,4),each=5)
```

Well, that saves space, but it's not obvious which condition is which.
Did you put control first or last? That could make a huge difference if you forget months down the line!

Factors are a way of splitting the difference here.
They store information as numbers, but associate those numbers with 'levels'.
We can make factors by using the ```as.factor()``` command, but usually they come about from reading data in.
```{r}
cond.fact = as.factor(cond.char)
```

When you print out the factor conditions, it looks like the character list.
But at the bottom, you see something that says 'Levels: Control DrugA DrugB DrugC'.
If you want to get the levels of a factor, you can use the levels() command:
```{r}
levels(cond.fact)
```

You shouldn't worry too much about factors beyond this.
This is the default way categorical data is read into R.
Most of the time, they will act just like character strings when you are doing analysis.
However, sometimes this means we will need to use special commands to deal with factors.
We'll deal with those commands as we come to them.

Factors Self Assessment
-----------------------

Does it make sense to add factors together? 

What about adding numbers to factors? 

What does R do if you try?