In Class Assignment 2
=====================

Vectors
-------

You can also store vectors of information, which contains a number of data points.
You build vectors with the ```c()``` function, like so:

`c(1,2,3,4)`

You can store these vectors in a variable

```{r}
a = c(1,2,3)
```

You can even combine two vectors with c()
```{r}
b = c(4,5,6)
c(a,b)
```

You can edit a vector with the following command

`fix(a)`

You can only store one type of data in a vector.
If you try to mix types, it will force the data to get along.
In the case below, it will make the numbers into characters.

```{r}
c(1,2,'a')
```

Let's say you want to make a vectors of the number 1 to 10 - you could type this out the long way

`c(1,2,3,4,5,6,7,8,9,10)`

But that's a lot of work - there's an easier way.
The x:y operator returns a vector of values from x to y.

`1:10`

You can also use the seq() function to do the same thing.

`seq(1,10, by = 1)`

Note the use of 'by = 1',
sometimes functions will have a lot of different inputs, some of which don't need to be used.
By 'naming' the arguments, you make sure you're giving input correctly (and making your code easier to read).
In this case, 'by' doesn't need to be used at all (the default is 1 if nothing is entered):
```{r}
seq(1,10)
```

So why use 'seq' instead of ':'? Because it's more flexible.
Let's say we wanted all of the one digit odd numbers...
That's just the numbers from 1 to 9, incrementing by 2 each time:
```{r}
seq(1,9,by = 2)
```



Or let's say we want all of the numbers divisible by 5 from 0 to 100
```{r}
seq(0,100,by = 5)
```

Or we want to count backwards by 2 from 20
```{r}
seq(20,0,by = -2)
```

But let's say we don't want to count, we just want a vector with 1 repeated 10 times.
There's a function for that - rep()
```{r}
rep(1,times = 10)
```

You can also repeat vectors
```{r}
rep(c(1,2),times = 5)
```

Note that this gives you 1,2,1,2,...
If we want 1,1,1,1,1,2,2,2,2,2 there's the 'each' argument
```{r}
rep(c(1,2),each = 5)
```

You might still be curious why the [1] is at the start of all output.
It's because R is a 'vectorized' language - all variables are stored in vectors
Even single numbers are vectors of length 1.
The [1] tells you where in the vector you are starting on that line.

Sometimes you might have a longer vector - this helps to orient where in the vector you are.
In the example below, note that if the second line starts with [18], that line starts with the 18th entry in the vector.
```{r}
1:100
```

Now let's assign a few vectors to play around with
```{r}
v.1 = 1:5
v.2 = seq(2,10, by = 2)
v.3 = c(3,7,9,1)
```

If you want to access individual vector items, you can do that too if you know the number.
For instance, to get the first element of v.1, you would type:
```{r}
v.1[1]
```

For the third, type:
```{r}
v.1[3]
```



Note that to do this, you put square brackets after the vector, and give the index of the value you want to get.
Indices start at 1 in R, and go through the length of the vector.
If you try to access a number with an index that's out of bounds, it will return an NA value:
```{r}
v.1[10]
```

There are also a few basic functions you should know on vectors:

First, the ```length()``` function tells you how long a vector is (e.g., how many data points it holds)
```{r}
length(v.1)
```

Next, you can add everything in a vector together with sum()
```{r}
v.2
sum(v.2)
2 + 4 + 6 + 8 + 10
```

You can also find the average value in a vector with mean()
```{r}
mean(v.2)
```

Or find the lowest value with min() or highest with max()
```{r}
min(v.2)
max(v.2)
```

You can also sort a list from smallest to largest with sort()
```{r}
v.3
sort(v.3)
```

Or sort from largest to smallest by telling the function that decreasing = TRUE
```{r}
sort(v.3, decreasing = TRUE)
```

If you have two vectors of the same length, you can add all of the elements together in a piecewise function with +
```{r}
v.1
v.2
v.2 + v.1
```

Other mathematical operators work the same way:
```{r}
v.2 - v.1
v.2 * v.1
v.2 / v.1
```

You can also add/subtract/multiply/etc. a single number to each entry in the vector
```{r}
v.1 + 2
v.1 * 2
```

Or test each element of a vector - this returns a vector of logicals
```{r}
v.1 >= 3
```

Want to know how many of the vector elements pass the conditional? Use sum()!
```{r}
sum(v.1 >= 3)
```

It's also possible to do arithmetic on vectors of unequal lengths - but THIS IS BAD!
R will 'recycle' the shorter vector, and repeat it so that it matches the longer one
```{r}
v.3 + c(1,2)
```

This will be the same as:
```{r}
v.3 + c(1,2,1,2)
```

I'm going to reiterate: *THIS IS BAD!*

In the rare case that you actually want to recycle vectors, you should use the rep() function.
For instance:
```{r}
v.3 + rep(c(1,2),times = 2)
```

This makes it easier to see what is going on and thus makes it more difficult for unwanted errors to slip in.
So don't let R recycle your vectors - ever!

Vectors and slicing
-------------------

Recall in the last lesson that you can create vectors with the c() command and access individual items by index
```{r}
junk.vector = c(2,6,4,10,8)
junk.vector[3] # Should be 4 - the item in the third position
```

But lets say you wanted a larger slice of the vector - perhaps the first three values.
You can get these in one swoop using the range 1:3
```{r}
junk.vector[1:3] # 2, 6, 4
```

Or the last three items in the vector
```{r}
junk.vector[3:5] # 4, 10, 8
```

But if we don't know exactly how long the vector is, we using the length function.
This tells it to get back all items from index (length-2) to the end (length)
```{r}
junk.vector[(length(junk.vector)-2):length(junk.vector)] # 4, 10, 8
```

Note the parentheses... if you don't use them, this will solve as length - (2:length) which becomes 3:0...
```{r}
junk.vector[length(junk.vector)-2:length(junk.vector)]
junk.vector[3:0] # 4, 6, 2 (position 0 is ignored)
```


Likewise, you can get all items EXCEPT certain indices using negative values.
For instance, if you want everything but the first entry, you would write
```{r}
junk.vector[-1] # 6, 4, 10, 8
```

And if you want everything except entries 2 through 4, you would write:
```{r}
junk.vector[-2:-4] # 2, 8
```

You can also 'name' vector entries using the names() command.
By default all names are null... you can check this:
```{r}
names(junk.vector)
```

But you can then set the names by assigning a vector of names to that command:
```{r}
names(junk.vector) = c("Alan","Barbara","Cathryn","Dylan","Eric")
```

Now the vector prints with names
```{r}
junk.vector
```

You can also check the names directly again
```{r}
names(junk.vector)
```

And now you can pull from the vector by name
```{r}
junk.vector["Dylan"]
```

Or multiple names
```{r}
junk.vector[c('Alan','Eric')]
```

Or erase the names
```{r}
names(junk.vector) = NULL
junk.vector
```

You can also get only values that satisfy a conditional by placing that conditional in the brackets.
For instance, say you wanted only the values in ```junk.vector``` that are greater than 5:
```{r}
junk.vector[junk.vector > 5]
```

This works because you get a vector of logicals out of ```junk.vector > 5```

```{r}
junk.vector > 5 # F T F T T
```

And when you access a vector with a vector of equal length logicals, it returns just the TRUE ones
```{r}
junk.vector[c(FALSE,TRUE,FALSE,TRUE,TRUE)]
```

This works just as well for conditionals from other vectors
```{r}
cond.vector = c("Red","Blue","Red","Blue","Green")
cond.vector == "Red" # T F T F F
junk.vector[cond.vector == "Red"]
```

On top of just indexing data in a vector, you might want to change it.
If you want to change a single index, this is easy - you just assign to that index
```{r}
junk.vector
junk.vector[3] = 7
junk.vector
```

You can also assign to a range of indices
```{r}
junk.vector[2:4] = c(8,10,2)
junk.vector
```

Note that you must be careful that the length of the sub-vector you are overwriting is the same length as what you feed it.
Recycling will occur otherwise, and I hope I don't have you remind you how bad that is.
Seriously... don't do it.

You might also want to assign a value to everything that satisfies a given condition.
This might happen if you want to enact a ceiling/floor, or code for error data.
For instance, what if we want anything greater than 9 in junk.vector to be set to 9 (a ceiling)?
```{r}
junk.vector[junk.vector > 9] = 9
junk.vector
```

Ungraded Self Assessment
------------------------
Make a vector of the classes you've taken (in order), including 851.
```{r}
courses <- c("CSE 891", "ZOL801", "ZOL896", "ZOL832", "ZOL849", "MMG801", "ZOL827", "STT814", "ZOL851", "BLD870", "NSC820", "ZOL999")
```

Make a new vector from the old that only includes graduate level classes (500 level and above):
```{r}
courses_above_850 
```

Slice the original vector to exclude the classes you are currently in.
```{r}
#Your Code Here
```
